What is an Ancestral Recombination Graph?

### 2022-02-02 paper.tex

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMCRA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

The limitations of the Griffiths ARG encoding directly stem from its
rooting in the coalescent with recombination stochastic process.
Indeed, these limitations arise because it is such a \emph{good}
mathematical encoding, providing sufficient information to
define the realised ancestry under the assumptions of the model,
and no more.

@@

Due to the computational expense of sampling ``little ARGs", state-of-the-art
Monte Carlo methods typically rely on cheaper, approximate models.
The most prominent examples are \texttt{ARGWeaver} \citep{rasmussen2014genome}
and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
discretisation approximation, but scales to dozens of mammal-like genomes.
The \texttt{ClonalOrigin} method \citep{didelot2010inference,
medinaaguayo2020speeding} uses the \texttt{ClonalFrame}
model and data structure, though it is not as scalable as \texttt{ARGWeaver}.

>>>>

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced by Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of this process is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''

@@

The limitations of the Griffiths ARG encoding are a direct consequence
of its derivation from the coalescent with recombination stochastic process.
Indeed, these limitations arise because it is such a \emph{good}
mathematical encoding, providing sufficient information to
define the realised ancestry under the assumptions of the model,
and no more.

@@

Due to the computational expense of sampling ``little ARGs", state-of-the-art
Monte Carlo methods typically rely on cheaper, approximate models.
The most prominent examples are \texttt{ARGWeaver} \citep{rasmussen2014genome}
and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
discretisation approximation, but scales to dozens of mammal-like genomes.
The \texttt{ClonalOrigin} method \citep{didelot2010inference,
medina2020speeding} uses the \texttt{ClonalFrame}
model and data structure, though it is not as scalable as \texttt{ARGWeaver}.
